The hidden manifold distance for functional manifold data

### 2019-06-24 index.Rmd
modularizing real data analysis

---
title: "Project Description"
output:
  html_document:
    df_print: paged
---

@@

## Abstract

Title: Pairwise geodesic distance estimation for functional data residing near a nonlinear manifold.

Abstract: Certain functional data of interest reside in non-linear spaces; the class of probability density functions provides one such example. Since these spaces are not vector spaces, even basic operations such as addition and subtraction require special consideration and treatment. In this work we set ourselves the task of calculating the geodesic distance between two functions living near a non-linear manifold due to noise. This setting cannot be tackled by classic manifold learning techniques which require data to live exactly on a manifold. Good estimation of the geodesic distance has implications for many downstream tasks in functional data analysis, including functional clustering, and nonparametric functional regression.

## Introduction

The classical setup in FDA assumes we observe a sample of $n$ independant realizations $X_1,\ldots,X_n$ of a random variable $X$ that takes value in the Hilbert space $L^2([a,b],\R)$. In the classical approach to analysing such data, we first obtain a functional version $\tilde X_1,\ldots,\tilde X_n$ of the raw data (by either smoothing or using PACE), next perform FPCA to reduce the dimension of the data, and then proceed by employing multivariate techniques.

However, it might very well happen that the function $X$ belongs to $\M \subset L^2([a,b],\R)$ a low-dimensional nonlinear manifold (gives examples and references). If it is indeed the case, it may no longer be appropriate to use $L^2$ distance between two curves as a measure of distance. Scenario 2 in [sim_functional_data](sim_functional_data.html) is one such example. One may want to consider using some geodesic distance which takes into account the intrinsic structure of $\M$.

@@

## Preliminaries

@@

### Basic differential geometry
This is take from a paper on [metric learning](http://proceedings.mlr.press/v32/linb14.pdf).
Let $(\M,g)$ be a $d$-dimensional Riemannian manifold, where $g$ is a Riemannian metric tensor on $\M$. This metric tensor can be used to assign a metric on the manifold. For each point $p$ on the manifold, the Riemannian metric tensor $g$ has an inner product $g_p$ on the tangent space $T_p \M$. The norm of a tangent vector $v \in T_p \M$ is defined as $$||v|| = \sqrt{g_p(v,v)}$$
If $\gamma: [a,b] \subset \R \to \M$ is a smooth curve, its length is then defined as
$$ l(\gamma) := \int_a^b || \frac{\,d\gamma}{\,d t}(t) || \,dt $$
The geodesic distance between two points $p,q$ on the manifold $\M$, based on this metric tensor $g$, is defined as
$$ d_g(p,q):=\inf \{l(\gamma): \gamma:[a,b] \to \M \text{ piecewise smooth}, \gamma(a) = p, \gamma(b) = q]  \}.$$

### IsoGeo

@@

## Proposed method for estimating geodesic distances

@@

## Simulaton Study

We perform a simulation study to ascertain the efficacy of our method for estimating pairwise geodesic distances for discretely-observed noisy functional data. ???Describe simulation scenarios in sim_functional_data.R???

# Comparison to other methods

@@

# Assessment of geodesic distance estimates
We introduce a new metric for assessing the quality of a pairwise geodesic distance estimator. This is the ROC curve with $\epsilon$ on the $x$-axis and degree to which near-$\epsilon$ isometry holds, i.e.\ the percentage of estimated pairwise distances between $1-\epsilon$ and $1+\epsilon$ of the truth pairwise distance.

## TODOs

* Related to tuning parameter
    + Choose the number of neighbors $N$ in IsoGeo.
    + Choose projection dimension for avoiding curse of dimensionality in performing density ridge scms routine
    + Choose the bandwidth $h$ in scms.
* Related to spartanSim.R simulations results
    + Why do some Monte Carlo simulations fail?
    + Is P-isomap always choosing zero penalty automatically?
    + Shouldn’t the smoothing method be doing better? Thought the smooth was close to the true curve.
* Next time we submit spartanSim.R, make sure to hold dataset constant for different projection dimensions
* Read Marie's suggested paper on [robust ISOMAP](https://arxiv.org/abs/1306.3373)
* Revisit Mueller manifold examples. Need to write down some differential equations to solve the geodesic minimisation problem?

>>>>

---
title: "The hidden manifold distance for functional manifold data"
output:
  html_document:
    df_print: paged
    number_sections: true
---

@@

# Abstract

Certain functional data of interest reside in non-linear spaces; the class of probability density functions provides one such example. Since these spaces are not vector spaces, even basic operations such as addition and subtraction require special consideration and treatment. In this work we set ourselves the task of calculating the geodesic distance between two functions living near a non-linear manifold due to noise. This setting cannot be tackled by classic manifold learning techniques which require data to live exactly on a manifold. Good estimation of the geodesic distance has implications for many downstream tasks in functional data analysis, including functional clustering, and nonparametric functional regression.

# Introduction

The classical setup in FDA assumes we observe a sample of $n$ independant realizations $X_1,\ldots,X_n$ of a random variable $X$ that takes value in the Hilbert space $L^2([a,b],\R)$. However, it might very well happen that the function $X$ belongs to $\M \subset L^2([a,b],\R)$ a low-dimensional nonlinear manifold (gives examples and references). If it is indeed the case, it may no longer be appropriate to use $L^2$ distance between two curves as a measure of distance. 
<!-- Scenario 2 in [sim_functional_data](sim_functional_data.html) is one such example.  -->
One may want to consider using some geodesic distance which takes into account the intrinsic structure of $\M$.

@@

## Setting
Describe preprocessing of discretely observed functional data: In the classical approach to analysing such data, we first obtain a functional version $\tilde X_1,\ldots,\tilde X_n$ of the raw data (by either smoothing or using PACE), next perform FPCA to reduce the dimension of the data, and then proceed by employing multivariate techniques.

What type of noise structures are allowed/considered? What type of time sampling frequency is assumed? Do not need time-aligned curves!

# Preliminaries

@@

## Basic differential geometry
This is take from a paper on [metric learning](http://proceedings.mlr.press/v32/linb14.pdf).
Let $(\M,g)$ be a $d$-dimensional Riemannian manifold, where $g$ is a Riemannian metric tensor on $\M$. This metric tensor can be used to assign a metric on the manifold. For each point $p$ on the manifold, the Riemannian metric tensor $g$ has an inner product $g_p$ on the tangent space $T_p \M$. The norm of a tangent vector $v \in T_p \M$ is defined as $$||v|| = \sqrt{g_p(v,v)}$$
If $\gamma: [a,b] \subset \R \to \M$ is a smooth curve, its length is then defined as
$$ l(\gamma) := \int_a^b || \frac{\,d\gamma}{\,d t}(t) || \,dt $$
The geodesic distance between two points $p,q$ on the manifold $\M$, based on this metric tensor $g$, is defined as
$$ d_g(p,q):=\inf \{l(\gamma): \gamma:[a,b] \to \M \text{ piecewise smooth}, \gamma(a) = p, \gamma(b) = q]  \}.$$

## IsoGeo

@@

# Proposed method for estimating geodesic distances

@@

## Selection of tuning parameters

# Simulation study

We perform a simulation study to ascertain the efficacy of our method for estimating pairwise geodesic distances for discretely-observed noisy functional data.

## Alternative estimators of geodesic distance

@@

## Simulation scenarios

## Performance metrics
We introduce a new metric for assessing the quality of a pairwise geodesic distance estimator. This is the ROC curve with $\epsilon$ on the $x$-axis and degree to which near-$\epsilon$ isometry holds, i.e.\ the percentage of estimated pairwise distances between $1-\epsilon$ and $1+\epsilon$ of the truth pairwise distance.

## Results

# Study on "misspecification"

Suppose we take scenario 1 where the geodesic distance coincides with the $L_2$ distance. Then what is the cost of employing our distance estimator compared to smoothing the data and doing numerical integration to find the $L_2$ distance.

# Downstream tasks

In this section, we explore whether our geodesic distance estimator has benefits for downstream analysis task. There are many tasks we could consider here, but we will focus on distanced-based functional clustering and distance-based functional classification. It must be noted that while curve alignment, also known as curve registration, is necessarily performed as a preprocessing technique prior to clustering and classification, our geodesic distance estimator allows one to forsake this step.

In both tasks below, we compare our method to using $L_2$ distance, possibly weighted, and with curve registration already accomplished.

## Distance-based functional clustering

Describe the $k$-medoids clustering method.

### Simulated data

###  Real data

## Distance-based functional classification

For simplicity, assume the task is bianry classification. Associated to each functional object $x$ is a binary $y$ indicating class membership.
Consider the classifier proposed in Ferraty and Vieu (2003,2006) which is a functional version of the Nadaraya-Watson kernel estimator of class membership probabilities:
$$
\hat p(y = 0 | x) \frac{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] 1(y_i = 0) }{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] }
$$

### Simulated data

###  Real data
A few examples of the datasets used by functional classification papers

* Wheat, rainfall and phoneme in "Achieving near-perfect classification for functional data"
* Berkeley growth curves in Chen et. al 2014, Biometrics.
* Tecator and phoneme in technometrics paper.
